---
title: "Geospatial Blog Post"
description: "Creating a Function for Coastal Species Habitat"
author:
  - name: Erica Dale
    url: http://ericamarie9016.githubt.io
    affiliation: MEDS
    affiliation-url: http://ucsb-meds.github.io
date: 2022-12-02
format:
  html:
    code-fold: true
    code-summary: "Show the code"
code-overflow: wrap
code-block-bg: true
code-block-border-left: "#6B5A75"
categories: [MEDS, Geospatial, R, West Coast, Oysters]
citation: 
  url: http://ericamarie9016.github.io/2022-12-02-geospatial
#bibliography: references.bib
# image: 
# draft: TRUE
---

## Introduction

This project explores the suitable habitats for species along the Western Coast, beginning with an Oyster and then creating a function to explore any species. The parameters to explore are depth and sea surface temperature. This could futher be expanded with more variables, including change of temperature within the water column. These regions are broken down to five Exclusive Economic Zones (EEZ) along the West Coast of the US.

This could be applied for restoration efforts to determine where habitat is suitable for any oceanic species.\

### Data

#### Sea Surface Temperature

Average annual sea surface temperature (SST) from the years 2008 to 2012 are used to characterize the average sea surface temperature within the region. Originally generated from [NOAA's 5km Daily Global Satellite Sea Surface Temperature Anomaly v3.1](https://coralreefwatch.noaa.gov/product/5km/index_5km_ssta.php).

#### Bathymetry

To characterize the depth of the ocean: [General Bathymetric Chart of the Oceans (GEBCO)](https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area).[^1]

[^1]: GEBCO Compilation Group (2022) GEBCO_2022 Grid (<doi:10.5285/e0f0bb80-ab44-2739-e053-6c86abc0289c>).

#### Exclusive Economic Zones

Exclusive Economic Zones off of the west coast of US: [Marineregions.org](https://www.marineregions.org/eez.php).

## Data Import

```{r include=TRUE, warning=FALSE, messages=FALSE}
library(terra)
library(here)
library(sf)
library(tidyverse)
library(tmap)
library(ggpubr)    # Combines ggplots
```

```{r}
Regions <- read_sf("~/MEDS/Fall-EDS223-Geospatial/data/data_4/wc_regions_clean.shp")
st_crs(Regions)       # WGS84 EPSG 4326

Annual2008 <- rast("~/MEDS/Fall-EDS223-Geospatial/data/data_4/average_annual_sst_2008.tif")
Annual2009 <- rast("~/MEDS/Fall-EDS223-Geospatial/data/data_4/average_annual_sst_2009.tif")
Annual2010 <- rast("~/MEDS/Fall-EDS223-Geospatial/data/data_4/average_annual_sst_2010.tif")
Annual2011 <- rast("~/MEDS/Fall-EDS223-Geospatial/data/data_4/average_annual_sst_2011.tif")
Annual2012 <- rast("~/MEDS/Fall-EDS223-Geospatial/data/data_4/average_annual_sst_2012.tif")
SSTAnnual <- c(Annual2008, Annual2009, Annual2010, Annual2011, Annual2012)     # stack the annual
st_crs(SSTAnnual)    # WGS84, EPSG 9122

Depth <- rast("~/MEDS/Fall-EDS223-Geospatial/data/data_4/depth.tif")
st_crs(Depth)        # WGS84, EPSG 4326



#### Reproject ###

SSTAnnual <- project(SSTAnnual, Depth)
st_crs(SSTAnnual)     # WGS84, EPSG 4326
```

#### Data Prep

```{r include=TRUE}
### Find mean SST ###

# Collapse down layers to one SST mean
SSTMean <- mean(SSTAnnual)
SSTMean <- (SSTMean$mean - 273.15)
plot(SSTMean)
```

```{r}
### Crop depth raster to match extent of SST raster ###
# crop()

# Already matched due to projection method
#Depth         # extent      : -132, -114, 29, 50
#SSTAnnual     # extent      : -132, -114, 29, 50
```

```{r}
### Match resolutions ###

# Already matched due to projection method
#Depth        #resolution  : 0.004166667, 0.004166667  (x, y)
#SSTAnnual    #resolution  : 0.004166667, 0.004166667  (x, y)

## IS THIS A PROBLEM SINCE IT WASN'T NEAREST NEIGHBOR??

#Check
SSTDepth <- c(Depth, SSTMean)
```

#### Find suitable locations

```{r include=TRUE}
# sea surface temperature: 11-30°C = 284.15 - 303.15
# depth: 0-70 meters below sea level

### Set up reclassification matrices ###
SSTrcl <- matrix(c(-Inf, 11, NA,
                   11, 30, 1,   
                   30, Inf, NA),
                     ncol = 3, byrow = TRUE)
OysterSST <- classify(SSTMean, rcl = SSTrcl)
plot(OysterSST, col = "blue")


Depthrcl <- matrix(c(-Inf, -70, NA,
                     -70, 0, 1,
                     0, Inf, NA),
                     ncol = 3, byrow = TRUE)
OysterDepth <- classify(Depth, rcl = Depthrcl)
plot(OysterDepth, col = "blue")



### Overlay # Find areas that satisfy BOTH

fun_mult = function(x,y){return(x*y)}     # Function to multiply layers

OystersHabitat <- lapp(c(OysterSST, OysterDepth), fun_mult)
plot(OystersHabitat, col = "blue")
```

#### Determine the most suitable EEZ

```{r include=TRUE}
##### Turn Regions data into raster
Regions$rgn <- as.factor(Regions$rgn)
RegionsRast <- rasterize(vect(Regions), OystersHabitat, field = "rgn")
plot(RegionsRast)


##### Select suitable cells in EEZ
ROmask <- mask(RegionsRast, OystersHabitat)


##### Find area of grid cells
expanse(ROmask, unit = "km")          # 11099.3 km

# layer value      area
# [1,]     1     0 3864.5258
# [2,]     1     1  150.4956
# [3,]     1     2 1089.7404
# [4,]     1     3 3467.7569
# [5,]     1     4 2526.7837

##### Find total suitable area within each EEZ
# Attempt 1
#ROZone <- zonal(OystersHabitat, RegionsRast, fun="sum", na.rm = TRUE, as.raster = TRUE)
#plot(ROZone)


OysterZone <- expanse(ROmask, unit = "km", byValue = TRUE)
OysterZone

# Attempt 2
# zonal(OystersHabitat, RegionsRast, fun="sum", na.rm = TRUE)


##### Find percentage of each zone
  
# Join suitable areas by region onto EEZ vector data (Regions)
# spatial join

# math
OysterRegion <- cbind(OysterZone, Regions)
OysterRegion <- OysterRegion |> 
  mutate(percent = (area/area_km2)*100)

#    percent
# 1 2.14702963
# 2 0.09155413
# 3 0.53751081
# 4 1.67637234
# 5 3.77705171

print(paste("Central California has 3864.5258 km2,  2.14% of the EEZ. Northern California has 150.4956 km2, .09% of the EEZ. Orgeon has 1089.7404 km2, with .53%. Southern California has 3467.7569 km2, with 1.67%. Washington has 2526.7837 with 3.77%"))
```

#### Visualization

```{r include=TRUE}
# Total suitable area by region
tm_shape(OysterRegion) +
  tm_polygons("area", palette = "YlGn") +
  tm_text(text = "rgn",
          size = .7) +
  tm_layout(main.title = "Suitable Oyster Area",
            main.title.size = 1.2,
            main.title.position = "center") +
  tm_layout(legend.position = c("left", "bottom"))
  

# Percent suitable area by region
tm_shape(OysterRegion) +
  tm_polygons("percent", palette = "YlGn") +
  tm_text(text = "rgn",
          size = .7) +
  tm_layout(main.title = "Suitable Oyster Percentage",
            main.title.size = 1.2,
            main.title.position = "center") +
  tm_layout(legend.position = c("left", "bottom"))
```

#### Creating a Function

```{r}
Species_Habitat <- function(species_name, temp_min, temp_max, depth_max, depth_min){
  
  # RECLASSIFICATION #
  SSTrcl <- matrix(c(-Inf, temp_min, NA,
                   temp_min, temp_max, 1,
                   temp_max, Inf, NA),
                   ncol = 3, byrow = TRUE)
  Depthrcl <- matrix(c(-Inf, depth_max, NA,
                     depth_max, depth_min, 1,
                     depth_min, Inf, NA),
                     ncol = 3, byrow = TRUE)  
  
  # APPLY RECLASS TO BOTH #
  SpeciesSST <- classify(SSTMean, rcl = SSTrcl)
  SpeciesDepth <- classify(Depth, rcl = Depthrcl)
  SpeciesHabitat <- lapp(c(SpeciesSST, SpeciesDepth), fun_mult)
  
  # # EEZ #
  # ##### Select suitable cells in EEZ
  Regions$rgn <- as.factor(Regions$rgn)
  RegionsSpecRast <- rasterize(vect(Regions), SpeciesHabitat, field = "rgn")
  RSpecmask <- mask(RegionsSpecRast, SpeciesHabitat)
  # Checked to work to here

  # ##### Find total suitable area within each EEZ
  SpeciesZone <- expanse(RSpecmask, unit = "km", byValue = TRUE)
  SpeciesZone # Error on row number matching???

  # ##### Find percentage of each zone
  SpeciesRegion <- cbind(SpeciesZone, Regions)
  SpeciesRegion <- SpeciesRegion |>
     mutate(percent = (area/area_km2)*100)
  
  # Total suitable area by region
  SpecAreaMap <- tm_shape(SpeciesRegion) +
  tm_polygons("area", palette = "YlGn") +
  tm_text(text = "rgn",
          size = .7) +
  tm_layout(main.title = print(paste0(species_name,  " Suitable Area")),
            main.title.size = 1.2,
            main.title.position = "center") +
  tm_layout(legend.position = c("left", "bottom"))
  

  # Percent suitable area by region
  SpecPercMap <- tm_shape(SpeciesRegion) +
  tm_polygons("percent", palette = "YlGn") +
  tm_text(text = "rgn",
          size = .7) +
  tm_layout(main.title = print(paste0(species_name, " Suitable Percentage")),
            main.title.size = 1.2,
            main.title.position = "center") +
  tm_layout(legend.position = c("left", "bottom"))
  
  tmap_arrange(SpecAreaMap, SpecPercMap)
  
}


Species_Habitat("Dolphin", 11, 30, -70, 0)
```
